Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-07-27 paint-contour-tree.tex
section 8.2 added

>>>>

\newcommand{\buff}{\mathcal{B}}
\newcommand{\loss}{\lambda}
\subsection{Pruning Away Buffered Sub-shrubs}
Let $\cU$ be a shrub in $\cL \setminus \cL'$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.

% In the following, any connected subgraph of a shrub will be called a \emph{sub-shrub} (i.e. it may not be a proper subtree).
% 
% \begin{definition}
%  Let $\cU$ denote a shrub of $\cL \setminus \cL'$.  
%  Then a sub-shrub $\buff$ of $\cU$ is called \emph{buffered} if the following holds:
%  \begin{enumerate}
%   \item $\sum_{p\in \buff} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff} |p|$
%   \item $\buff$ is minimal, i.e.\ there is no sub-shrub of $\buff$ for which the above inequality holds.  
%  \end{enumerate}
% \end{definition}

Ultimately our goal in this subsection is to prune away subtrees from $\cU$ which can be paid for by charging to their tall support chains (similarly to supported short paths). The remaining portion of the shrub will then be handled in a later subsection. We now formally define this pruning process, which is quite intuitive if one thinks of $\cU$ as an actual physical shrub.

\begin{definition}
  Let $\cU$ denote a shrub of $\cL \setminus \cL'$. A \emph{pruning} $\cU'$ of $\cU$ is any subgraph defined by some number of rounds of the following iterative process.  Initially set $\cU' = \cU$.  In each iteration select some subtree $\cU'_r$ rooted at a node $r\in \cU'$ and set $\cU' = \cU'\setminus \{\cU'_r\}$.  
\end{definition}

The above definition does not specify which subtrees are pruned, which we now formally define.

\begin{definition}\label{def:buff}
 Let $\cU'$ be a pruning of a shrub of $\cU$ of $\cL \setminus \cL'$. Then a rooted subtree $\buff$ of $\cU'$ is called a \emph{buffered sub-shrub} if the following conditions hold:
  \begin{enumerate}
  \item $\sum_{p\in \buff} |\loss(p)| \geq \frac{1}{100} \sum_{p\in \buff} |p|$
  \item No rooted subtree of $\buff$ satisfies the above inequality (other than $\buff$ itself). 
 \end{enumerate}
\end{definition}

% \begin{definition}
%  Let $\cU$ denote a shrub of $\cL \setminus \cL'$. We define an interative  procedure which ``prunes'' $\cU$.  Specifically, let $\cU'$ denote the current remaining portion of the shrub, so initially, $\cU' = \cU$.  A single iteration consists of the following: Visit the nodes in $\cU'$ in a reverse topological order (i.e. leaves first).  Let $q$ be the current node being considered, and let $\cU'_q$ be the subtree rooted at $q$.  We say $\cU'_q$ is buffered if the following condition holds
%  \[
%   \sum_{p\in \cU'_q} \loss(p) \geq \frac{1}{100} \sum_{p\in \cU'_q} |p|
%  \]
%  An iteration stops when we reach a node $q$ defining a buffered subtree, at which point we set $\cU' = \cU'\setminus{\cU'_q}$.
% \end{definition}

Ultimately we will prune buffered subtrees from $\cU$ by starting at the leaves and working towards the root, and hence the second condition of buffered will trivially be satisfied.

We require the following lemma so that we can charge to the root in a buffered subtree.

\begin{lemma}
 Let $p$ be any node in a buffered sub-shrub $\buff$, and let $d$ be any descendant node of $p$ in $\buff$.  Then for any $v \in d$, $|H_v| = O(|H_p|)$.
\end{lemma}
\begin{proof}
 Missing.  \Lem{cu-root} will be rephrased to handle this slightly more general case (i.e. here $p$ may not be the root of $\cU$). Note the proof will not 
 have to be changed as \Lem{geometric} is the proof which already implies general case.
\end{proof}

\begin{lemma}
 Let $\buff$ be a buffered sub-shrub.  Then $\sum_{p\in \buff}\sum_{v\in p} \log |H_v| = O(\sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| )$
\end{lemma}
\begin{proof}
 ????  Ignore this ``proof'', it is just a graveyard of observations. ????

 Let $r$ be the root of $\buff$. First observe that $\loss(r) \geq |r|/100$.  Suppose otherwise, then in order for the first condition of \Def{buff} to be satisfied for $\buff$, the rooted subtree of some child of $r$, call it $\buff'$, must satisfy $\sum_{p\in \buff'} \loss(p) \geq \frac{1}{100} \sum_{p\in \buff'} |p|$.  However, this would contradict the second condition of \Def{buff}.

 Consider any node $q\in \buff$.  Then for any vertex $v\in \loss(q)$, we have $|H_v| = \Omega(|H_q|)$, since $q$ is unsupported and hence by definition less than $|H_q|/100$ vertices of $H_q$ appear in $\loss(q)$.

 \begin{align*}
  \sum_{p\in \buff}\sum_{v\in p} \log |H_v| &=  O\left( \sum_{p\in \buff}\sum_{v\in p} \log |H_p| \right)   &\mbox{definition of short} \\
  &= O\left( \sum_{p\in \buff} |p| \log |H_p| \right)   &\mbox{} \\
  &= ????   &\mbox{buffered is not sufficient to get to next line} \\
  &= O\left( \sum_{p\in \buff} |\loss(p)| \log |H_p| \right)   &\mbox{} \\
  &= O\left( \sum_{p\in \buff}\sum_{v\in \loss(p)} \log |H_v| \right)
 \end{align*}
\end{proof}

Let $p$ and $q$ be two distinct paths in $\cL \setminus \cL'$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.

\begin{corollary}
 Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cL \setminus \cL'$. Then, 
 $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$
\end{corollary}
